\textbf{ Interpretable Gaussian Processes for Starspots }

### 2021-01-20 tex/ms.tex
tweaks

The code was designed to maximize the speed and numerical stability of the
computation. Although most of the computations of the expectation integrals
(see Appendix~\ref{sec:integrals}) involve many layers of nested sums over
all spherical harmonic coefficients, these may be expressed as high-dimensional
tensor products, which can be evaluated extremely efficiently on modern
hardware. Many of the expressions can also be either pre-computed or
computed recursively. To maximize the speed of the algorithm, the code is
implemented in hybrid \cpp/\Python using the just-in-time compilation
capability of the \theano package. Since all equations derived here
have closed form expressions, these can be differentiated in a straightforward
and numerically stable manner, enabling the computation of backpropagated
gradients within \theano. As such, \starryprocess is designed to work
out-of-the box with \theano-based inference tools such as \pymcthree
for NUTS/HMC or ADVI sampling.

\begin{figure}[t!]
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/speed.pdf}
        \oscaption{speed}{%
            Evaluation time in seconds for a single log-likelihood evaluation
            as a function of the number of points $K$ in each light curve
            when conditioning on a value of the inclination (blue) and when
            marginalizing over the inclination (orange). At $l_\mathrm{max} = 15$,
            computation of the covariance matrix of the GP takes about 10ms
            on a 2018 MacBook Pro. The dashed line shows the asymptotic scaling
            of the algorithm, which is due to the Cholesky factorization and
            solve operations.
            \label{fig:speed}
        }
    \end{centering}
\end{figure}

@@

Because our expression for the GP covariance has an exact, closed-form
solution as a function of the spot parameters, it can be computed
extremely efficiently: a typical likelihood evaluation on a dataset
consisting of $K \sim 1{,}000$ points takes about 10ms on a modern
laptop. Our algorithm is implemented in the open-source, user-friendly
\Python package \starryprocess, which is \textsf{pip}-installable and available on
\href{https://github.com/rodluger/starry_process}{GitHub}.
The algorithm is implemented in a combination of \cpp and \Python,
linked using the \theano package. Because our GP covariance has an exact
representation, so too do its derivatives. We therefore implement
backpropagated derivatives with respect to all input parameters for
out-of-the-box usage with gradient-based inference and optimization
tools such as Hamiltionain Monte Carlo (HMC), autodifferentation
variational inference (ADVI), and gradient-based nested sampling.

We devoted a large portion of this paper to testing the algorithm on a variety
of synthetic datasets, showing that it is a well-calibrated and in most
cases unbiased estimator for star spot properties. Below we list the main
results presented in this paper:

@@

    \item \textbf{In general, an ensemble of at least $M \sim 50$ light curves
              is needed to place meaningful constraints on the spot properties}.
          This estimate is based on ensemble analyses of light curves with
          $K = 1{,}000$ cadence each and per-cadence precision of one part
          per thousand. Lower quality, shorter baseline observations, or
          datasets contaminated by outliers will in general require larger values
          of $M$ for the same constraining power. The presence of strong limb
          darkening also degrades the information content of light curves, in
          which case an ensemble of hundreds or even a thousand light curves
          is recommended.

@@

    \item \textbf{Relative single-band photometry cannot uniquely constrain
              the spot contrast, the total number of spots, or the total spot
              coverage}. These two quantities
          are perfectly degnerate with each other in relative single-band photometry.
          Knowledge of the
          unspotted photospheric brightness (or temperature) can break this
          degeneracy, so analyses of multi-band light curves can in principle
          infer the total spottiness of a stellar surface.

@@

          rotation, will be the subject of the next paper in this series.

@@

>>>>

The code was designed to maximize the speed and numerical stability of the
computation. Although most of the computations of the expectation integrals
(see Appendix~\ref{sec:integrals}) involve many layers of nested sums over
all spherical harmonic coefficients, these may be expressed as high-dimensional
tensor products, which can be evaluated efficiently on modern
hardware. Many of the expressions can also be either pre-computed or
computed recursively. To maximize the speed of the algorithm, the code is
implemented in hybrid \cpp/\Python using the just-in-time compilation
capability of the \theano package. Since all equations derived here
have closed form expressions, these can be differentiated in a straightforward
and numerically stable manner, enabling the computation of backpropagated
gradients within \theano. As such, \starryprocess is designed to work
out-of-the box with \theano-based inference tools such as \pymcthree
for NUTS/HMC or ADVI sampling.

\begin{figure}[t!]
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/speed.pdf}
        \oscaption{speed}{%
            Evaluation time in seconds for a single log-likelihood evaluation
            as a function of the number of points $K$ in each light curve
            when conditioning on a value of the inclination (blue) and when
            marginalizing over the inclination (orange). At $l_\mathrm{max} = 15$,
            computation of the covariance matrix of the GP takes about 20ms
            on a 2018 MacBook Pro. The dashed line shows the asymptotic scaling
            of the algorithm, which is due to the Cholesky factorization and
            solve operations.
            \label{fig:speed}
        }
    \end{centering}
\end{figure}

@@

Because our expression for the GP covariance has an exact, closed-form
solution as a function of the spot parameters, it can be computed
efficiently: a typical likelihood evaluation on a dataset
consisting of $K \sim 1{,}000$ points takes about 20ms on a modern
laptop. Our algorithm is implemented in the open-source, user-friendly
\Python package \starryprocess, which is \textsf{pip}-installable and available on
\href{https://github.com/rodluger/starry_process}{GitHub}.
The algorithm is implemented in a combination of \cpp and \Python,
linked using the \theano package. Because our GP covariance has an exact
representation, so too do its derivatives. We therefore implement
backpropagated derivatives with respect to all input parameters for
out-of-the-box usage with gradient-based inference and optimization
tools such as Hamiltionain Monte Carlo (HMC), autodifferentation
variational inference (ADVI), and gradient-based nested sampling.

We devoted a large portion of this paper to testing the algorithm on a variety
of synthetic datasets, showing that it is a well-calibrated and in most
cases unbiased estimator for star spot properties. Below we list the main
results concerning our GP:

@@

@@

@@

          rotation, will be the subject of a future paper in this series.

@@

In this paper we also demonstrated and discussed four points that apply more
broadly to stellar light curve variability modeling in general:

\begin{enumerate}
    \item \textbf{Typically, an ensemble of at least $M \sim 50$ light curves
              is needed to place meaningful constraints on starspot properties}.
          This estimate is based on ensemble analyses of light curves with
          $K = 1{,}000$ cadence each and per-cadence precision of one part
          per thousand. Lower quality, shorter baseline observations, or
          datasets contaminated by outliers will in general require larger values
          of $M$ for the same constraining power. The presence of strong limb
          darkening also degrades the information content of light curves, in
          which case an ensemble of hundreds or even a thousand light curves
          is recommended.
    \item \textbf{Limb darkening breaks degeneracies}. The presence of limb darkening
          changes the structure of the null space. Even small variations in the
          limb darkening coefficients among stars in an ensemble can effectively
          increase the information content of their light curves when analyzed
          jointly. However, this is both a blessing and a curse: it is very
          important to accurately model limb darkening to infer starspot
          properties without bias.
    \item \textbf{Relative single-band photometry cannot uniquely constrain
              the spot contrast, the total number of spots, or the total spot
              coverage}. We showed that these two quantities
          are perfectly degenerate with each other in relative single-band photometry.
          Knowledge of the
          unspotted photospheric brightness (or temperature) can break this
          degeneracy, so analyses of multi-band light curves can in principle
          infer the total spottiness of a stellar surface.
    \item \textbf{Care should be taken when analyzing normalized light curves}.
        We showed that the widely-adopted practice of normalizing light curves
        to their mean, median, or maximum changes the covariance structure of
        the data in potentially complex ways. In addition to changing the total
        variance of the data, this normalization step introduces non-stationarity
        in the covariance, which can lead to biases particularly when
        modeling light curves of stars with very long rotation periods.
\end{enumerate}
